---
title: "Maturity-at-age"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: FALSE
    code_folding: show
    number_sections: TRUE
---

# SET-UP

```{r markdown, echo=F}
library(knitr)
opts_chunk$set(echo = T, collapse = T, fig.align = "center", fig.width = 9, fig.height = 6,warning = FALSE, message = FALSE)
options(width = 300)
```    

```{r settings, message = F}
source('0.0_settings.R')
```

# Input data
## Read
```{r read, message = F}
new <- FALSE
if(new){
    bio <- get.bio(species='maquereau',user=imlp.user,password=imlp.pass)
    f <- paste0('Rdata/bio_',Sys.Date(),'.Rdata')
    save(bio,file=f)
}else{
    df <- file.info(list.files("Rdata/", full.names = T,pattern="bio_"))
    f <- rownames(df)[which.max(df$mtime)]
    load(f)
}
print(f)
maa.gear <- read.table(paste0(dir.dat,'maa_group_gear.txt'),header=T)
maa.region <- read.table(paste0(dir.dat,'maa_group_region.txt'),header=T)
```

## Clean
```{r clean, message = F}
# subset
bio.mat <- bio[,c('year','nafo','gear','sex','month','length.frozen','agef','matur')]
names(bio.mat)[c(6:7)] <- c('length','age')
nrow(bio.mat)

# what months?
bio.mat <- bio.mat[bio.mat$month %in% c(6:7),]   # was done before
# bio.mat <- bio.mat[bio.mat$month %in% c(5:11),]    # less smooth results, depsite larger amount of data and small effect of month on maa

# remove NAs
bio.mat <- bio.mat[!is.na(bio.mat$matur) & !is.na(bio.mat$age) & !is.na(bio.mat$year),]
nrow(bio.mat)

# remove fish with ages that cannot be trusted (wrong length: see 3.0_caa))
bio.mat <- bio.mat[bio.mat$age<18,]      
bio.mat <- ddply(bio.mat,c('age'),transform,outlier=outlier(length,coef=3))  
bio.mat[(bio.mat$age==0 & bio.mat$length>300),'outlier'] <- TRUE     
bio.mat[is.na(bio.mat$outlier),'outlier'] <- FALSE
bio.mat <- bio.mat[bio.mat$outlier==FALSE,]
bio.mat$outlier <- NULL
nrow(bio.mat)

### transform
# group gears and regions for sens test
bio.mat$region <- maa.region[match(bio.mat$nafo,maa.region$nafo),'region']
table(bio.mat$region,useNA = 'always')

bio.mat$gear.group <- maa.gear[match(bio.mat$gear,maa.gear$gear.cat),'gear.group']
table(bio.mat$gear.group,useNA = 'always')

# age: 10 is plus group
bio.mat[bio.mat$age>10,'age'] <- 10

# mature vs immature
bio.mat$mat <- ifelse(bio.mat$matur<3,0,1) # based on maturity stage, not sex (F, I, M)

# correct maturity stage of age 0 fish. If caught in months 8-12 can impossibly be mature because only some months old.
bio.mat[bio.mat$mat==1 & bio.mat$age==0 & bio.mat$month>7,'mat'] <- 0

# proportion mature at age
prop.mat <- ddply(bio.mat,c('year','age'),summarise,
               prop.immat=length(mat[mat==0])/length(mat),
               prop.mat=length(mat[mat==1])/length(mat),
               n=length(mat))
```

## Tables {.tabset}
### numbers
```{r table numbers}
kable(t(table(bio.mat$mat,bio.mat$year)))
```

## Plots {.tabset}

### age vs maturity curve
```{r plot age_mat,fig.width=18}
ggplot(prop.mat,aes(x=age,y=prop.mat))+
    geom_line(size=1)+
    geom_point()+
    geom_text(aes(label=n),size=3,hjust=0,vjust=0)+
    facet_wrap(~year)+
    scale_color_viridis_c()
```

### age vs maturity tile
```{r plot age_mat_tile,fig.width=10,fig.height=5}
ggplot(prop.mat,aes(x=year,y=age,fill=prop.mat))+
    geom_tile()+
    geom_text(aes(label=round(prop.mat,2)),size=2)+
    scale_fill_viridis_c()+
    scale_x_continuous(expand=c(0,0))+
    scale_y_continuous(expand=c(0,0))
```

### age vs maturity n
```{r plot age_mat_tilen,fig.width=10,fig.height=5}
ggplot(prop.mat,aes(x=year,y=age,fill=n))+
    geom_tile()+
    geom_text(aes(label=n),size=2)+
    scale_fill_viridis_c(direction = -1)+
    scale_x_continuous(expand=c(0,0))+
    scale_y_continuous(expand=c(0,0))
```

# Calculations

```{r calcs, message = F}
# exclude 1973
bio.mat <- bio.mat[bio.mat$year!=1973,]
prop.mat <- prop.mat[prop.mat$year!=1973,]

# run models
mods <- lapply(unique(bio.mat$year), function(x) glm(mat~age,data=bio.mat[bio.mat$year==x,],family=binomial(logit))) # warning not a problem

# predictions 
df <- data.frame(age=seq(0,10,0.1))
preds <- lapply(mods, function(x) cbind(df,predict(x,df,type="response",se.fit=TRUE)))           # predictions
preds <- lapply(preds, function(x) cbind(x, pup= x$fit+1.96*x$se.fit))     # add upper bound
preds <- lapply(preds, function(x) cbind(x, plow= x$fit-1.96*x$se.fit))    # add lower bound

names(preds) <- unique(bio.mat$year)
preds <- bind_rows(preds,.id='year')

maa <- preds[preds$age %in% 1:10,]
rownames(maa) <- 1:nrow(maa)
maa$fit <- round(maa$fit,2)

# test:
# mod <- glm(mat~age+as.factor(year)+as.factor(gear.group)+as.factor(region)+month,data=bio.mat,family=binomial) # though there are interactions
# df2 <- data.frame(expand.grid(age=seq(0,10,0.1),year=as.factor(unique(bio.mat$year))),month=6,region='sGSL',gear.group='Lines')
# preds2 <- cbind(df2,predict(mod,df2,type="response",se.fit=TRUE))          # predictions
# preds2$pup <- preds2$fit+1.96*preds2$se.fit     # add upper bound
# preds2$plow <- preds2$fit-1.96*preds2$se.fit    # add lower bound

```

## Plots {.tabset}

### fit
```{r plot pred,fig.width=18,fig.height=24 }
ggplot(preds)+
    geom_rug(data=bio.mat[bio.mat$mat==0,],aes(x=age,y=mat),sides='b', position = "jitter",col='grey') +
    geom_rug(data=bio.mat[bio.mat$mat==1,],aes(x=age,y=mat),sides='t', position = "jitter",col='grey') +
    geom_ribbon(aes(ymin=plow,ymax=pup,x=age),fill='red',alpha=0.5)+
    geom_line(aes(x=age,y=fit),col='red',size=1)+
    geom_point(data=prop.mat,aes(x=age,y=prop.mat))+
    scale_y_continuous(limits=c(0,1),expand=c(0,0))+
    labs(x='Age',y='Proportion mature')+
    facet_wrap(~year)
```

### maa
```{r plot maa}
ggplot(maa,aes(x=as.numeric(year),y=fit))+
    geom_point(aes(col=as.factor(age)))+
    geom_line(aes(col=as.factor(age)),size=1)+
    labs(x='Year',y='Proportion mature',col='Age')+
    scale_color_viridis_d()+
    theme(legend.position = 'none')
```

### maa ci
```{r plot maaci}
ggplot(maa,aes(x=as.numeric(year),y=fit))+
    geom_ribbon(aes(ymin=pmax(0,plow),ymax=pmin(1,pup),fill=as.factor(age)),alpha=0.5)+
    geom_point(aes(col=as.factor(age)))+
    geom_line(aes(col=as.factor(age)),size=1)+
    labs(x='Year',y='Proportion mature',col='Age')+
    scale_color_viridis_d()+
    theme(legend.position = 'none')
```

# Sensitivity: age 0
What if there were actually had observations of age 0 (not selected by fishery) that are immature?
Conclusion: no significant difference.

```{r sens_padd0, message = F}
# add age 0
n <- 100 # number of new immature fish each year
new <- data.frame(year=unique(bio.mat$year),age=0,mat=0)  # all new age 0 for all years
bio.matsupp <- rbind(bio.mat[,which(names(bio.mat) %in% names(new))],new[rep(seq_len(nrow(new)), n), ])

# run models
modssupp <- lapply(unique(bio.matsupp$year), function(x) glm(mat~age,data=bio.matsupp[bio.matsupp$year==x,],family=binomial))

# predictions 
predssupp <- lapply(modssupp, function(x) cbind(df,predict(x,df,type="response",se.fit=TRUE)))           # predictions
predssupp <- lapply(predssupp, function(x) cbind(x, pup= x$fit+1.96*x$se.fit))     # add upper bound
predssupp <- lapply(predssupp, function(x) cbind(x, plow= x$fit-1.96*x$se.fit))    # add lower bound

names(predssupp) <- unique(bio.matsupp$year)
predssupp <- bind_rows(predssupp,.id='year')

maas <- predssupp[predssupp$age %in% 1:10,]
rownames(maas) <- 1:nrow(maas)

p1 <- ggplot(maas,aes(x=as.numeric(year),y=fit))+
    geom_ribbon(aes(ymin=plow,ymax=pup,fill=as.factor(age)),alpha=0.5)+
    geom_point(aes(col=as.factor(age)))+
    geom_line(aes(col=as.factor(age)),size=1)+
    labs(x='Year',y='Proportion mature',col='Age',title='Base')+
    scale_color_viridis_d()+
    theme(legend.position = 'none')+
    scale_y_continuous(limits = c(0,1.01),expand = c(0,0))
p2 <- ggplot(maas,aes(x=as.numeric(year),y=fit))+
    geom_ribbon(aes(ymin=plow,ymax=pup,fill=as.factor(age)),alpha=0.5)+
    geom_point(aes(col=as.factor(age)))+
    geom_line(aes(col=as.factor(age)),size=1)+
    labs(x='Year',y='Proportion mature',col='Age',title='Padded with age 0')+
    scale_color_viridis_d()+
    theme(legend.position = 'none')+
    scale_y_continuous(limits = c(0,1.01),expand = c(0,0))
grid.arrange(p1,p2)
```

# Comparison with before

## all years
```{r comp, fig.width=15}
repo <- "https://github.com/iml-mackerel/0.0_model/blob/master/"

ys <- c(2016,2018,2020)
maa.hist <- lapply(ys, function(x) read.ices(url(paste0(repo,'data/',x,'/mo.dat',"?raw=true"))))
names(maa.hist) <- ys 

maam <- lapply(names(maa.hist), function(x) reshape2::melt(as.matrix(maa.hist[[x]]),varnames=c('year','age'),value.name=x))
maa.comp <- Reduce(function(x, y) merge(x, y, all=TRUE), maam)

maa.new <- maa[,c('year','age','fit')]
names(maa.new)[3] <- '2022'

maa.comp <- merge(maa.comp,maa.new,all=TRUE)
maa.comp <- melt(maa.comp,id=c('year','age'))
ggplot(maa.comp,aes(x=as.numeric(year),y=value,col=variable))+
    geom_line(size=1)+
    facet_wrap(~age,scale='free_y')+
    labs(x='Year', y='Proportion mature',col='Assessment year')

```

# Fill and smooth (not used)

Fit Noel Cadigan's state-space model for waa.
1) filter out noise.
2) fill in gaps.

Fits an AR1 process in 3 directions (age, year, cohort). Available in catchR package.

Would need logit transformation to be usefull.. not log.
Unsure this makes much sense because annual model output.
Will not use this

## Fit model
Keep all years and observations
```{r maa_fit, message = F, fig.height = 11,fig.width = 10,results = 'hide',message=FALSE}
maa.in <- maa
maa.in[maa.in$fit==0 & !is.na(maa.in$fit),'fit'] <- 0.001  # to be able to take log and no CV of Inf
maa.in$cv <- with(maa.in,se.fit/fit)

all <- expand.grid(year=1968:max(maa$year), age=min(maa$age):max(maa$age))
maaf <- merge(all, maa.in, all.x = TRUE)
maaf$year <- as.numeric(maaf$year)
maaf$cv <- 0.6  # doesn't work with the mini numbers estimated

maa.fit <- armatrix.fit(year=maaf$year,age=maaf$age,x=maaf$fit,cv=maaf$cv,shrink.cv = 0)
maa.fit  # convergence ok

maa.fit$output$pred_exp <- pmin(maa.fit$output$pred_exp,1)
```

## Plots {.tabset}

### waa
```{r maa_smoothed, message = F, fig.height = 3,fig.width = 5,results = 'hide',message=FALSE}
armatrix.waa(maa.fit,ylab='Proportion mature')+scale_color_viridis_d()
```

# Fill and smooth simple

```{r maa_smooth, message = F}
maas <- maa[,c('year','age','fit')]
maas$year <- as.numeric(maas$year)

# set to one for older ages (once 1 reached cannot decrease anymore)
maas[maas$age>4,'fit'] <- 1

# add early years
av <- ddply(maas[maas$year %in% 1974:1979,],c('age'),summarise,fit=mean(fit))
toadd <- cbind(year=rep(1968:(min(maas$year)-1),each=length(unique(maas$age))),
               av[rep(seq(nrow(av)),6),])
maas <- rbind(toadd,maas)

# remove years with insufficeint data
thres <- 30
toremove <- ddply(prop.mat[prop.mat$age %in% 1:2,],c('year'),summarise,toremove=ifelse(sum(n)<30,T,F))
maas[maas$year %in% toremove[toremove$toremove,'year'],'fit'] <- NA

ggplot(maas,aes(x=as.numeric(year),y=fit))+
    geom_point(aes(col=as.factor(age)))+
    geom_line(aes(col=as.factor(age)),size=1)+
    labs(x='Year',y='Proportion mature',col='Age')+
    scale_color_viridis_d()+
    theme(legend.position = 'none')
```

## compare with 2020 smoothed
```{r compsmooth, fig.width=6,fig.height=6}
setwd('..')
y <- 2020
load(paste0('00.0_model/Rdata/',y,'/fit.Rdata'))
mo <- fit$data$propMat
mo <- melt(mo,varnames = c('year','age'),value.name = 'fit')
mo$source <- 'OLD'

source(paste0('00.0_model/R/smoothmatrix.R'))
sm <- smoothmatrix(dcast(maas,year~age,value.var = 'fit')[,-1],smooth = 0.5)
sm[sm>1] <- 1
sm[sm<0] <- 0
sm$year <- min(maas$year):max(maas$year)
sm <- melt(sm,id='year', variable.name = 'age',value.name = 'fit')
sm$source <- 'NEW'

sms <- rbind(mo,sm)
sms$age <- as.numeric(as.character(sms$age))
ggplot(sms,aes(x=year,y=fit,col=as.factor(age)))+
    geom_line(size=1)+
    facet_wrap(source~.,ncol=1)+
    labs(x='Year',y='Proportion mature',col='Age')+
    scale_y_continuous(limits=c(0,1.01),expand=c(0,0))+
    scale_x_continuous(expand = c(0,0))+
    scale_color_viridis_d()
```

# Save results
```{r waa_result, message = F}
s <- dcast(sm,year~age,value.var = 'fit')
s[,2:ncol(s)] <- round(s[,2:ncol(s)] ,3)
write.csv(s, file=paste0('csv/maa_',Sys.Date(),'_base_smooth0.5.csv'),row.names = FALSE)

p <- ggplot(sms[sms$source=='NEW',],aes(x=year,y=fit,col=as.factor(age)))+
    geom_line(size=1)+
    labs(x='Year',y='Proportion mature',col='Age')+
    scale_y_continuous(limits=c(0,1.01),expand=c(0,0))+
    scale_x_continuous(expand = c(0,0))+
    scale_color_viridis_d()
ggsave(filename = paste0('img/maa',Sys.Date(),'_base_smooth0.5.png'),plot = p,units = 'cm',height = 8,width = 14)

```